Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-24 paper/submission.tex
Improve wording and add some context

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. In the ideal case, {\color{blue}from} knowing the movement of one price, the movement of a second can be inferred.  In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

  The concept of an entirely peer-to-peer digital currency managed via a distributed ledger was described and applied by Nakamoto in 2008 \cite{nakamoto2008bitcoin}, who named the currency `Bitcoin'. The proposal and subsequent implementation captured the attention of technologists, economists, libertarians and futurists, and spawned numerous adaptations utilising the blockchain technology \cite{aste2017blockchain}, which have come to be known as cryptocurrencies. Trading in these cryptocurrencies has become widely available even to less sophisticated retail investors, and volumes have grown significantly as interest in the currencies has widened. The crypto market is characterised by high volatility which seems to reflect changes in the attitudes of investors. The usage of cryptocurrencies in the traditional economy remains limited, and it is reasonable to assume that prices are in part driven by speculative dynamics, separate to any utility as a medium of exchange or to any revenue-generating process. Therefore one should expect to observe a larger interplay between sentiment and prices in cryptocurrencies, compared to more fundamental-driven markets such as equities. We discuss existing literature that identifies signals between social sentiment and equity prices, and so hypothesise that investor sentiment on future cryptocurrency prices may be expected to feed into the short-term price movements via speculation, which may be detected using applied information-theoretic techniques. This paper tests this hypothesis.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price {\color{blue} returns depend} on prior values of both {\color{blue} price returns} and market sentiment, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the {\color{blue} change in} social sentiment at $t-k$ on the {\color{blue} returns} at time $t$.

@@

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on {\color{blue}Granger-Geweke causality} ({\color{blue}the} linear causality measure) and transfer entropy ({\color{blue}the} non-linear causality measure). In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear {\color{blue}Granger} causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect {\color{blue}statitical} causality in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence {\color{blue} the time-dependence required for statistical causality}.

@@

    \includegraphics[width=\linewidth]{images/confirming_logmap.pdf}
    \caption{Demonstration that the non-linear causal relationship in synthetic data generated from equations \ref{eq:coupled_map} and \ref{eq:g(x)} is detected only by the non-linear method. The plots are calculated from the mean over 10 realisations over 15,000 data points of the synthetic coupled logistic map process, with $\epsilon = 0.4$. Non-linear transfer entropy is calculated using a histogram of 6 classes per dimension, partitioned using marginal equiquantisation. The Z-score for each result is also plotted for both methods. In the direction $X \rightarrow Y$, the classical approach is unable to detect the known dependency of Y on X, whilst the information-theoretic approach correctly identifies this.  We note that for $\alpha>0.2$ the information-theoretic method detects information transfer in the other direction. This is observed to increase as $\alpha$ approaches $1$, and corresponds to an expected non-linear anticipatory signal.
    }
    \label{fig:CLM_confirmation}
    %\vspace{-35pt} % Add this to rightsize the float and let it fit on a page
  \end{figure*}
  %\clearpage

>>>>

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. {\color{blue} This may be considered as a field of price discovery, which represents a large and fundamental area of financial theory.} In the ideal case {with complete predictive information, \color{blue}from} knowing the movement of one price, the movement of a second can be inferred {\color{blue} exactly. In reality information is incomplete and held asymmetrically, and the ability to reliably obtain this and contribute to price discovery is often well rewarded}. In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

  The concept of an entirely peer-to-peer digital currency managed via a distributed ledger was described and applied by Nakamoto in 2008 \cite{nakamoto2008bitcoin}, who named the currency `Bitcoin'. The proposal and subsequent implementation captured the attention of technologists, economists, libertarians and futurists, and spawned numerous adaptations utilising the blockchain technology \cite{aste2017blockchain}, which have come to be known as cryptocurrencies. Trading in these cryptocurrencies has become widely available even to less sophisticated retail investors, and volumes have grown significantly as interest in the currencies has widened. The crypto market is characterised by high volatility which seems to reflect changes in the attitudes of investors. The usage of cryptocurrencies in the traditional economy remains limited, and it is reasonable to assume that prices are in part driven by speculative dynamics, separate to any utility as a medium of exchange or to any revenue-generating process. %Therefore one should expect to observe a larger interplay between sentiment and prices in cryptocurrencies, compared to more fundamental-driven markets such as equities. 
  {\color{blue} One factor which contributes to these dynamics is market sentiment, the estimation of which using social media data is an active area of research}.  We discuss existing literature that identifies signals between social sentiment {\color{blue} data} and equity prices, and so hypothesise that investor sentiment on future cryptocurrency prices may be expected to feed into the short-term price movements via speculation, which may be detected {\color{blue} through the usage of} information-theoretic techniques. This paper tests this hypothesis.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price {\color{blue} returns depend} on prior {\color{blue} returns and changes in market sentiment}, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the {\color{blue} change in} social sentiment at $t-k$ on the {\color{blue} returns} at time $t$.

@@

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on {\color{blue}Granger-Geweke causality} ({\color{blue}the} linear causality measure) and transfer entropy ({\color{blue}the} non-linear causality measure). In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear {\color{blue}Granger} causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect {\color{blue}statistical} causality in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence{\color{blue}, in expectation, the time-dependence required for there to be any statistical causality}.

@@

    \includegraphics[width=\linewidth]{images/confirming_logmap.pdf}
    \caption{Demonstration that the non-linear causal relationship in synthetic data generated from equations \ref{eq:coupled_map} and \ref{eq:g(x)} is {\color{blue}correctly} detected only by the non-linear method. The plots are calculated from the mean over 10 realisations over 15,000 data points of the synthetic coupled logistic map process, with $\epsilon = 0.4$. Non-linear transfer entropy is calculated using a histogram of 6 classes per dimension, partitioned using marginal equiquantisation. The Z-score for each result is also plotted for both methods. In the direction $X \rightarrow Y$, {\color{blue} the classical approach is unable to detect the known dependency of Y on X, whilst detecting spurious small signals. By contrast the information-theoretic approach correctly identifies the relationship}.  We note that for $\alpha>0.2$ the information-theoretic method detects information transfer in the other direction. This is observed to increase as $\alpha$ approaches $1$, and corresponds to an expected non-linear anticipatory signal.
    }
    \label{fig:CLM_confirmation}
    %\vspace{-35pt} % Add this to rightsize the float and let it fit on a page
  \end{figure*}
  %\clearpage
